On Adding Pattern Matching to Haskell-based Deeply Embedded Domain Specific Languages\\ \large{Not For Distribution}

### 2020-07-24 paper/paper.tex
Improve a type signature in Church encoding section
TODO about nested pattern match -> note in paper
Conal citation in 'Motivation' and more explaination of tail recursion transformation

Additionally, lambdas can also be difficult to implement in EDSLs. A major reason
for this is that control flow constructs which are typically \textit{outside} the EDSL,
such as pattern matching and tail recursion, can make it challenging to "look inside"
the lambda.
\begin{todo}
  Maybe add brief summary of relationship to lambdas?
\end{todo}

@@

A major benefit of the Haskell language is that it is frequently possible derive
a Haskell implementation from a specification, either partially or fully. When
it is not possible to fully derive an implementation from a specification, it is
often easier to check a Haskell implementation against a specification than it
is in many imperative languages due, in part, to the management of side effects
in Haskell. This is more difficult to do using an imperative language such as C,
as they provide no restrictions on where side effects can occur.

@@

\begin{todo}
  Cite examples where Haskell implementations have been derived specifications.
  Conal Elliott's latest two papers (as of May 2020),
  \cite{Elliott-2018-ad-icfp} and \cite{Elliott2019-convolution-extended}, would
  be good examples of this.

@@

\begin{todo}
  Should we use a single nested pattern match in the above block? Both
  are functional (and equivalent) from a technical perspective.
\end{todo}

@@

\begin{lstlisting}
church_mkPair
  :: a -> b -> (a -> b -> r) -> r
church_mkPair = \ x y f -> f x y
\end{lstlisting}

@@

\begin{lstlisting}
church_fst :: ((a -> b -> a) -> r) -> r
church_fst = \ p -> p (\ x y -> x)

church_snd :: ((a -> b -> b) -> r) -> r
church_snd = \ p -> p (\ x y -> y)
\end{lstlisting}

@@

\begin{lstlisting}
church_mkLeft
  :: a -> (a -> r) -> (b -> r) -> r
church_mkLeft = \ x f g -> f x

church_mkRight
  :: b -> (a -> r) -> (b -> r) -> r
church_mkRight = \ y f g -> g y
\end{lstlisting}

@@

\begin{lstlisting}
church_eitherElim
  :: (a -> b -> r) -> a -> b -> r
church_eitherElim = \ e f g -> e f g
\end{lstlisting}

@@

\begin{todo}
  Explicitly mention how the constructs in this last part
  relate to their counterparts in the main part of the paper.
\end{todo}

@@

\section{Representing tail recursion}
\begin{todo}
  Is the explanation in this section sufficient since we cite the Haskino paper,
  which describes how the tail recursion transformation was used in that project?
\end{todo}

Tail recursion is given a direct-style representation using a simple sum type,
using the technique described in~\cite{Grebe:2017:RSD:3136040.3136048}.

@@

We can take advantage of this view by transforming a tail recursive function
\ttt{f :: a -> b} into a function \ttt{f' :: a -> Iter b a}. \ttt{Iter b a} is a
type which can either correspond to an argument "update" (in the sense mentioned
previously) or a final result. This type is implemented as a sum of the types
\ttt{a} and \ttt{b}.

>>>>

Additionally, lambdas can also be difficult to implement in EDSLs. A major reason
for this is that control flow constructs which are typically \textit{outside} the EDSL,
such as pattern matching and tail recursion, can make it challenging to "look inside"
the lambda. When these control structures are reified within the EDSL, it is
easier to also reify lambdas, as you can simply apply them to a dummy argument
to access their body (with the dummy value substituted for the argument).
\begin{todo}
  Maybe add brief summary of relationship to lambdas?
\end{todo}

@@

A major benefit of the Haskell language is that it is frequently possible derive
a Haskell implementation from a specification, either partially or fully.
Examples of this technique include the work in \cite{Elliott-2018-ad-icfp} as
well as the work in \cite{Elliott2019-convolution-extended}.  When it is not
possible to fully derive an implementation from a specification, it is often
easier to check a Haskell implementation against a specification than it is in
many imperative languages due, in part, to the management of side effects in
Haskell. This is more difficult to do using an imperative language such as C, as
they provide no restrictions on where side effects can occur.

@@

\begin{todo}
  Cite examples where Haskell implementations have been derived specifications.  Conal Elliott's latest two papers (as of May 2020), \cite{Elliott-2018-ad-icfp} and \cite{Elliott2019-convolution-extended}, would be good examples of this. (This is now cited)

@@

Note that the above pattern matches can be written as a single nested pattern
match. Both forms compile down to the same Core representation.

@@

\begin{lstlisting}
church_mkPair
  :: a -> b -> (a -> b -> r) -> r
church_mkPair x y = \ f -> f x y
\end{lstlisting}

@@

\begin{lstlisting}
church_fst :: ((a -> b -> a) -> r) -> r
church_fst p = p (\ x y -> x)

church_snd :: ((a -> b -> b) -> r) -> r
church_snd p = p (\ x y -> y)
\end{lstlisting}

@@

\begin{lstlisting}
church_mkLeft
  :: a -> (a -> r) -> (b -> r) -> r
church_mkLeft x = \ f g -> f x

church_mkRight
  :: b -> (a -> r) -> (b -> r) -> r
church_mkRight y = \ f g -> g y
\end{lstlisting}

@@

\begin{lstlisting}
church_eitherElim
  :: ((a -> r) -> (b -> r) -> r) -> r
church_eitherElim e = \ f g -> e f g
\end{lstlisting}

@@

@@

\section{Representing tail recursion}
Tail recursion is given a direct-style representation using a simple sum type,
using the technique described in~\cite{Grebe:2017:RSD:3136040.3136048}.

@@

We can take advantage of this view by transforming a tail recursive function
\ttt{f :: a -> b} into a function \ttt{f' :: a -> Iter b a}. \ttt{Iter b a} is a
type which can either correspond to an argument "update" (in the sense mentioned
previously) or a final result. This type is implemented as a sum of the types
\ttt{a} and \ttt{b}. Recursive calls are transformed to \ttt{Step} applications
and non-recursive branches are wrapped in \ttt{Done} applications.
